---
title: "STA221"
author: "Neil Montgomery"
date: "Last edited: `r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output: 
  beamer_presentation:
    keep_tex: TRUE
    incremental: TRUE
    df_print: tibble
    fig_caption: FALSE
classoption: aspectratio=169
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE,
                      dev='pdf', fig.width=4, fig.asp=0.618, fig.align='center')
options(tibble.width=70, scipen = 999)
library(tidyverse)
library(readxl)
```

# applications of $\chi^2$ goodness-of-fit testing to two-way tables

## contingency tables

Recall the gas pipelines data:

```{r}
gas <- read_excel("pipeline.xlsx", col_types = rep("text", 4))
gas <- gas %>% 
  mutate(Leak=factor(Leak), Size=factor(Size, levels=c("1","1.5","1.75"), ordered=TRUE),
         Material=factor(Material), Pressure=factor(Pressure, levels=c("Low", "Med", "High"), ordered=TRUE))
gas
```

The (only?) suitable numerical summary for two categorical/factor variables at a time is a so-called contingency table, or two-way table.

## two-way table for "Leak" and "Pressure"

(Technically the two-way table doesn't include the Sum row and column.)

```{r, results='asis'}
library(xtable)
addtorow <- list()
addtorow$pos <- list(0, 0)
addtorow$command <- c("& \\multicolumn{3}{c}{Pressure} & \\\\\n",
"Leak & Low & Medium & High & Sum \\\\\n")

print(xtable(addmargins(table(gas$Leak, gas$Pressure)), digits = 0),
      comment = FALSE, add.to.row=addtorow, include.colnames=FALSE)
```

## the two questions

Question 1: do the rows (or columns) come from the *same* distribution?

\pause This question is answered using a *test of homogeneity*.

\pause Question 2: are the rows and columns *independent*?

\pause This question is answered using a *test of independence*.

\pause ***The mechanics of both tests are identical. Only the interpretation is (slightly) different.***

## two-way table again

Count version:

```{r, results='asis'}
library(xtable)
print(xtable(addmargins(table(gas$Leak, gas$Pressure)), digits = 0),
      comment = FALSE, add.to.row=addtorow, include.colnames=FALSE)
```

Proportion version. The six proportions at each combination of level of the two factor variables is the *joint distribution* of those two variables.

```{r, results='asis'}
library(xtable)
print(xtable(addmargins(prop.table(table(gas$Leak, gas$Pressure))), digits = 3),
      comment = FALSE, add.to.row=addtorow, include.colnames=FALSE)
```

## the marginal distrbutions

```{r, results='asis'}
library(xtable)
print(xtable(addmargins(prop.table(table(gas$Leak, gas$Pressure))), digits = 3),
      comment = FALSE, add.to.row=addtorow, include.colnames=FALSE)
```

The *marginal* distributions of `Pressure` and `Leak` are:
```{r, results='asis'}
print(xtable(t(as.table(addmargins(prop.table(table(gas$Leak, gas$Pressure)), 1)[3,])), digits= 3), comment=FALSE, include.rownames = FALSE)
print(xtable(t(as.table(addmargins(prop.table(table(gas$Leak, gas$Pressure)), 2)[,4])), digits= 3), comment=FALSE, include.rownames = FALSE)

```

## the conditional distributions

There are lots of conditional distributions. For example, the conditional distributions for the `Pressure` *given* `Leak` equals `No` and *given* `Leak` equals `Yes` are in the two rows of this table:

```{r, results='asis'}
print(xtable(prop.table(table(gas$Leak, gas$Pressure), margin = 1), digits=3),
      comment=FALSE)
```

The conditional distributions for `Leak` given `Pressure` is equal to, respectively, `Low`, `Med`, and `High`, are in these three columns:

```{r, results='asis'}
print(xtable(prop.table(table(gas$Leak, gas$Pressure), margin = 2), digits=3),
      comment=FALSE)
```

## diversion - if the marginal totals are *fixed*...

At some point there will be a "degrees of freedom" to consider, so let's do it now.

In all $\chi^2$ goodness-of-fit tests, **the overall sample sizes are considered to be *fixed***. This includes all the row and column totals in these two-way table analyses.

Consider the following table with fixed "marginal" totals. How many cells am I "free" to play around with?

\begin{table}[ht]
\centering
\begin{tabular}{r|r|r|r|r}
  \hline
  & \multicolumn{3}{c}{Factor A} & \\
 Factor B & 1 & 2 & 3 & Sum \\
 \hline
1 &  &  &  & 10 \\ \hline
2 &  &  &  & 20 \\ \hline
Sum & 5 & 10 & 15 & 30 \\ 
   \hline
\end{tabular}
\end{table}

\pause Answer: only **two**. With fixed marginal totals I have two "degrees of freedom". The formula is $(r-1)(c-1)$ when there are $r$ rows and $c$ columns.

## $\chi^2$ test of homegeneity

Do the rows (or columns) come from the *same* distribution? 

Specifically: do the rows (or columns) have the *same conditional disributions*?

\pause Let's compare the rows from before. They look pretty close, but not identical. 

```{r, results='asis'}
print(xtable(prop.table(table(gas$Leak, gas$Pressure), margin = 1), digits=3),
      comment=FALSE)
```

\pause The null hypothesis is "$H_0$: The rows have the same (conditional) distributions", and we keep all the marginal totals fixed.

## some technical details...

Let's get rid of the numbers from the tables and use some more general symbols.

The conditional distributions, which $H_0$ says are the same: 
\begin{table}[ht]
\centering
\begin{tabular}{rrrr}
  \hline
 & 1 & 2 & 3 \\ 
  \hline
1 & $p_{11}$ & $p_{12}$ & $p_{13}$ \\ 
2 & $p_{21}$ &$p_{22}$ & $p_{23}$  \\ 
   \hline
\end{tabular}
\end{table}

The counts, given *fixed* marginal totals:

\begin{table}[ht]
\centering
\begin{tabular}{rrrrr}
  \hline
 & 1 & 2 & 3 & Sum\\ 
  \hline
1 & $n_{11}$ & $n_{12}$ & $n_{13}$ & $n_{1.}$ \\ 
2 & $n_{21}$ &$n_{22}$ & $n_{23}$  & $n_{2.}$\\ 
Sum & $n_{.1}$ & $n_{.2}$ & $n_{.3}$ & $n_{..}$\\
   \hline
\end{tabular}
\end{table}

## the expected cell counts $E_{ij}$, under the null hypothesis

So we end up with:
\begin{align*}
E_{11} &= \frac{n_{1.}\cdot n_{.1}}{n_{..}}\\
\\
\onslide<2->{E_{ij} &= \frac{n_{i.}\cdot n_{.j}}{n_{..}}\\}
\end{align*}

\pause The counts we actually observe are called $O_{ij}$. We evaluate the deviation from $H_0$ using the formula:
$$\chi^2_{obs} = \sum\limits_{i,j} \frac{\left(O_{ij} - E_{ij}\right)^2}{E_{ij}}$$

## getting the p-value

No surprise that this sum of squares has a $\chi^2$ distributions. But with how many degrees of freedom?

\pause $(r-1)(c-1),$ where $r$ and $c$ are the numbers of rows and columns.

\pause In the example, the observed and expected cell counts are:

\begin{columns}
    \begin{column}{0.5\textwidth}
      \centering
```{r, results='asis'}
print(xtable(addmargins(with(gas, chisq.test(Leak, Pressure))$observed), digits=0), comment=FALSE)
```
    \end{column}
    \begin{column}{0.5\textwidth}
      \centering
```{r, results='asis'}
print(xtable(addmargins(with(gas, chisq.test(Leak, Pressure))$expected), digits=2), comment=FALSE)
```
    \end{column}
\end{columns}

## the full analysis

```{r}
(lp_hom <- with(gas, chisq.test(Leak, Pressure)))
```

$$P(\chi^2_2 \ge `r lp_hom$statistic`) = `r lp_hom$p.value`$$

There is no evidence against the null hypothesis. 

## example (Q23.14 from book)

All non-editorial publications from the NEJM were classified according to `Publication Year` and whether or not it contained a statistical analysis. 

```{r, results='asis'}
yrs <- factor(c("1978-79", "1989", "2004-05"))
sts <- factor(c("No Stats", "Stats"))
doctor_know_counts <- data.frame(cbind(expand.grid(Statistics=sts, 
                                            `Publication Year` = yrs), 
                 Counts=c(90, 242, 14, 101, 40, 271)), check.names = FALSE)
doctor_know <- xtabs(Counts ~ Statistics + `Publication Year`, doctor_know_counts)
print(xtable(addmargins(doctor_know), digits=0), comment=FALSE)
```

Question: "Has there been a change in the use of statistics?"

## example (Q23.14 from book)

Expected cell counts:

```{r}
doctor_know_chisq <- chisq.test(doctor_know)
round(addmargins(doctor_know_chisq$expected), 2)
```

\pause Results:

```{r}
doctor_know_chisq
```

## what's left to do?

* the test for independence,

* consider requirements and assumptions in order for these procedures to work,

* informal post-test assessments of the results
